\chapter{Sunflower Conjecture}

\section{Sunflower Lemma}

\begin{definition}
	[Sunflowers (Erd\"os-Rado, 1960's)]
	Take $f = \{ s_1 , \dots, s_m \}$ is a sunflower if all pairwise intersections are the same. i.e., 
	\begin{equation}
		S_i \cap S_j = S_k \cap S_e \quad \quad \text{if} \quad i \neq j, k \neq e.
	\end{equation}
	
	\textbf{\textit{Equivalently, }}
	$S_1, \dots, S_m$ is a sunflower if the ``core''
	\begin{equation}
		C = S_1 \cap S_2 \cap \dots \cap S_m
	\end{equation}
	are such that 
	\begin{equation}
		S_1 \text{\textbackslash} C, S_2 \text{\textbackslash} C, \dots, S_m \text{\textbackslash} C
	\end{equation}
	are disjoint
\end{definition}

\begin{lemma}
	[Sunflower Lemma (Erd\"os-Rado, 1962)] 
	Take $f = \{ S_1, \dots, S_m \} \subseteq [n]$, \footnote{
	This is a bit of abuse of notation. We really meant $S_i \subseteq [n], \forall i$ and $f$ is a set of such subsets. 
	}
	where $|S_i| \leq k, \forall i \in [m]$. Then $S$ contains a $r$-sunflower if 
	\begin{equation}
		m \geq (r - 1)^k \cdot k!
	\end{equation}
	
	\paragraph{English Explanation} This essentially says that any large collection of sets (large enough) must contain a sunflower. Interestingly enough, this expression does not depend on the size of universe, $n$. 
\end{lemma}

\begin{proof}
	The proof is an induction on $k$. 
	\paragraph{Base Case} $k = 1$. Then $S_i, S_j$ are disjoint and $m > (r - 1) \implies m \geq r$. So all of $f$ is a sunflower with empty core.
	\paragraph{Inductive Step} There are two cases to consider. Suppose the family $f$ had $ \geq r$ disjoint sets, then we are done. Otherwise, we choose some maximal collection of disjoint sets
	\begin{equation}
		S_{i1}, S_{i2}, ..., S_{il} \quad \quad \text{where} \quad l \leq r - 1
	\end{equation}
	This means that we cannot add another set that is disjoint from these. In particular, every other set must intersect with one of these sets. Now, call 
	\begin{equation}
		A \triangleq S_{i1} \cup S_{i2} \cup \dots \cup S_{il}
	\end{equation} 
	then the total number of elements in $A$ is bounded by 
	\begin{align}
		|A| 
		&\leq k \cdot l  \\
		&\leq k \cdot (r - 1) 
	\end{align}
	Now, every set $S_i$ must intersect $A$. Then, some element $a \in A$ must occur in $ \geq \alpha = m / |A|$ many sets. 
	\begin{align}
		\alpha 
		&\geq \frac{m}{k\cdot(r - 1)} \\
		&\geq \frac{(r - 1)^k \cdot k!}{k \cdot (r - 1)} \\
		&= (r - 1)^{k - 1} \cdot (k - 1)! \\
		\implies &m \geq (r - 1)^k \cdot k!
	\end{align}
	This concludes the proof. \qed
\end{proof}

\begin{proposition}
	[Erd\"os-Rado Conjecture]
	For every $r$, there is some constant $c_r$, such that if $m > c_r^k$, then it contains a $r$-sunflower. 
\end{proposition}

\begin{theorem}
	[ALWZ, 2020]
	There exists a constant $C$ such that if $f = \{ S_1, \dots, S_m \} \subseteq [n]$ and $|S_i| \leq k$, and if $m > (c \cdot r \cdot \log k ) ^k$, then $f$ contains a $r$-sunflower. 
\end{theorem}

\begin{definition}
	[Link of a Set System]
	Take $f = \{ S_1, \dots, S_m \} \subseteq [n]$. Suppose $I \subseteq [n]$, then define the link of $I$ in $f$ as 
	\begin{equation}
		f_i = \{ S_i : S_i \supseteq I \} \quad \quad \overline {f_I} = \{ S_i \setminus I : S_i \supseteq I \}
	\end{equation}
\end{definition}

\begin{definition}
	[Spread]
	A set system $f = \{ S_1, S_2, \dots, S_m \} \subseteq [n]$, where each $|S_i| \leq k$, is $s$-spread if 
	\begin{itemize}
		\item $|f| \geq s^k$, and 
		\item $\forall I \subseteq [n], |f_I| \leq s^{k - |I|} \cdot |f|$
	\end{itemize}
\end{definition}

\begin{lemma}
	[Main Lemma, ALWZ20]
	If $f$ is $s$-spread for 
	\begin{equation}
		s = c \cdot \log ( r k ) \log \log (r k)
	\end{equation}
	then $f$ contains $r$ disjoint sets. 
\end{lemma}

\begin{proposition}
	Main Lemma implies new sunflower lemma bounds. 	
\end{proposition}

\begin{proof}
	Suppose we have 
	\begin{equation}
		|f| \geq C(r, k)^k = s^k = (c \cdot \log ( r k ) \log \log (r k)) ^k
	\end{equation}
	If there exists a $I$ such that $|f_I| > s^{-|I|}$, then we are done. We see
	\begin{align}
		|f_I| 
		&\geq s^{-|I|} (c \cdot \log ( r k ) \log \log (r k))^k \\
		&= s^{k - |I|} \\
		&\geq \left( 
			c \cdot \log (r (k - |I|)) \cdot \log \log (r \cdot (k - |I|))
		\right)^{k - |I|}
	\end{align}
	so we rely on induction for this proof. If $\forall I$, $|f_I| < s^{- |I|} \cdot |f|$. Then, $f$ is $s$-spread ad by main lemma $f$ contains $r$ disjoint sets. \qed
\end{proof}

\section{Robust Sunflower Lemma}
\begin{lemma}
	[Robust Sunflower Lemma (RSL)]
	Consider $f$ as $s-spread$, and $w$ is a random subset of $[n]$ of size $\lfloor \frac{n}{r} \rfloor$. Then, 
	\begin{equation}
		Pr_{w} [\exists j, S_j \subseteq W] \geq 1 - \frac{1}{2r}
	\end{equation}
\end{lemma}

It is natural to wonder why we need this different form of the Sunflower Conjecture. In fact, this more robust RSL implies the Main lemma. 
\begin{proof}
	
\end{proof}


\begin{proposition}
	RSL $\implies$ MSL. 
\end{proposition}
This stronger form of the conjecture comes in handy when we are doing induction, we can have a more powerful hypothesis in the induction step. 

\begin{proof}
	Intuitively, consider the system to be a ``spread'' one, as then we can sample random patches $W$'s of the total space $[n]$, each of size $\lfloor \frac{n}{r} \rfloor$ elements. We expect that a patch covers some set $S_i$ with a reasonable chance ($(1 - \frac{1}{2r})$ stated in theorem). If this is the case, then if we take $r$ patches, we immediately know
	\begin{align}
		\text{RSL} \implies 
		\begin{cases}
			Pr[\exists j, S_j \subseteq W_1] = Pr[E_1] \geq 1 - \frac{1}{2r} \\
			Pr[\exists j, S_j \subseteq W_2] = Pr[E_2] \geq 1 - \frac{1}{2r} \\
			\quad \quad \vdots \\
			Pr[\exists j, S_j \subseteq W_r] = Pr[E_r] \geq 1 - \frac{1}{2r} 
		\end{cases}
	\end{align}
	Then, we can take the union bound of events to bound $Pr$[everything above happens at the same time] as 
	\begin{align}
		Pr[E_1 \cup E_2 \cup \dots \cup E_r] 
		&\leq Pr[E_1] + Pr_[E_2] + \dots + Pr[E_r] \\
		&= r \cdot \frac{1}{2r} \\
		&= \frac{1}{2} \\
		&> 0
	\end{align}
	So there $\exists$ $r$ disjoint sets. \qed
\end{proof}




















































